Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-03 application.tex
almost done with version 2

\section{Application to protoplanetary disks}\label{application} 
As a first application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one can consider the energy equation with
radiative cooling, 
%To connect our model with a realistic
%disk, we consider a more physically-motivated form of the energy
%equation with radiative cooling 
\begin{align}\label{real_energy}
%\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
  c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight, $c_v$ is the heat capacity at constant
volume, 
% In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists.

One should now model in detail each of the terms on the right
hand side (RHS) of Eq. \ref{real_energy}, linearize them, then solve
the resulting linear eigenvalue problem. Such an approach is beyond
the scope of this paper. Here, we simply model the RHS of
Eq. \ref{real_energy} such that it can be represented by a 
thermal relaxation time $t_c$ for the disk temperature. 
%Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
%the 
To do so, we first consider separately perturbations
with radial lengthscales $l\equiv 2\pi/k_x\gg l_\mathrm{rad}$ and 
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

\subsection{Radiative diffusion}
Let us first consider perturbations with lengthscales $l\gg
l_\mathrm{rad}$. In this case, we assume  
temperature fluctuations are smoothed out by radiative diffusion. We
mode this in linear theory by writing the linearized RHS of
Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%     c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
flucations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
% the radiative cooling is given by {\bf (reference?)}
% \begin{align}
% \nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
% \end{align}   
% where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
% heating term 
% \begin{align}
%   Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
% \end{align}
% Then Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
%   \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right), 
% \end{align}
% where $c_v$ is the heat capacity at constant volume. When linearized, the last term on the RHS becomes
% \begin{align}
%   -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
%     \frac{P}{\rho}\delta\rho\right). 
% \end{align}
% By comparing this expression with the linearized version of the energy
% equation used in our calculations (Eq. \ref{full_energy1}), we identify 
% \begin{align}\label{tc_newton_cool} 
%   t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
% \end{align}
% as the thermal relaxation timescale for optically thin cooling. 
\begin{align}
t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
 % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
%where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
%$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
%and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the 
% In order to utilize our linear results, which assumes $t_c$ is a
% constant, we need to make further 
% approximations. Specifically, we consider vertically-isothermal disks
% and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
% H$, where $\Sigma$ is the surface density at the fiducial
% radius of interest. For example, the dimensionless diffusion
% coefficient becomes 
% \begin{align}\label{diff_coeff_dimensionless}
%   \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
%     c_v\Omega_k}. 
% \end{align}
% Furthermore, we will choose a temperature-dependent opacity model for
% simplicity, so that $\kappa_d=\kappa_d(T)$.
dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{1}{\hat{\eta}}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that the thermal relaxation time depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane. %The latter is due to the density-dependence in the diffusion time. 
%is then a constant for vertically-isothermal disks, 
%and we can directly apply previous results.

@@

% The physical thermal relaxation timescale $\beta$ is height-dependent,
% whilst for analytical discussion we assumed it is
% constant. Nevertheless,
It is instructive to compare the above thermal timescales
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the fiducial disk model,    
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
% becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ and $\gamma=1.4$. We set the scaling $\hat{T}=\hat{\Sigma}=1$
and consider three opacity scales: $\hat{\kappa}_d=0.1, \,1 $ and $10$
at two heights: $z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in
each panel.

@@

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but raises also
them. In fact, lowering the opacity by an order of magnitude renders
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This makes
disk stable to the fundametal mode. Raising the opacity, as shown in
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the fundamental mode is
restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it develop in the inner disk, would only occur on very small
scales. It is not clear if such modes are robust (see below).

@@

\subsection{VSI in the MMSN}
We have solved the linear problem in the MMSN with the scale and
height dependent thermal relaxation function defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $1\lesssim \khat \lesssim 30$ we either only find the
fundamental mode, or it is the most unstable. The associated 
growth rates are also insensitive to $z_\mathrm{max}$. However, for
large $\khat$ and/or $z_\mathrm{max}$, e.g. $\khat=50$ or 
$z_\mathrm{max}=10H$, we also found modes with shorter growth times 
than those shown in Fig. \ref{mmsn_overall}, but these disturbances are 
concentrated near the vertical boundaries
\citep{nelson13,mcnally14}. Such modes may not be robust because    
they are affected by boundary conditions. Furthermore, modes with   
large $\khat$ are likely subject to viscous decay in a real disk.

We therefore take Fig. \ref{mmsn_overall} as a representation of the
VSI in the standard MMSN. We see that growth timescales are $\sim
30$---40 orbits, and the VSI operates from a few AU to tens of
AU. Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically thick.

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ only for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with
significantly reduced growth rate (as reflected in
Fig. \ref{mmsn_overall}).

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the growth rates are much
smaller than the case with $\beta < \beta_\mathrm{crit}$ for all $z$.

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one can consider the energy equation with
radiative cooling, 
\begin{align}\label{real_energy}
%\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
  c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
 \end{align}
%where the temperature $T$ is defined through $P=\mathcal{R}\rho
% T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
% the mean molecular weight, 
where $c_v$ is the heat capacity at constant
volume, 
% In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term such that an equilibrium solution exists.

One should now model in detail each of the terms on the RHS of
Eq. \ref{real_energy}, linearize them, then solve 
the resulting linear eigenvalue problem. Such an approach is beyond 
the scope of this paper. Here, we simply model the RHS of 
Eq. \ref{real_energy} such that it can be represented by a 
thermal relaxation time $t_c$ for the disk temperature. 
%Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
%the 
To do so, we first consider separately perturbations
with radial lengthscales $l\gg l_\mathrm{rad}$ and 
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume  
temperature fluctuations are smoothed out by radiative diffusion. We
mode this in linear theory by writing the linearized RHS of
Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%     c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
flucations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
% the radiative cooling is given by {\bf (reference?)}
% \begin{align}
% \nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
% \end{align}   
% where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
% heating term 
% \begin{align}
%   Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
% \end{align}
% Then Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
%   \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right), 
% \end{align}
% where $c_v$ is the heat capacity at constant volume. When linearized, the last term on the RHS becomes
% \begin{align}
%   -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
%     \frac{P}{\rho}\delta\rho\right). 
% \end{align}
% By comparing this expression with the linearized version of the energy
% equation used in our calculations (Eq. \ref{full_energy1}), we identify 
% \begin{align}\label{tc_newton_cool} 
%   t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
% \end{align}
% as the thermal relaxation timescale for optically thin cooling. 
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
% where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
% $\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
% and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the 
% In order to utilize our linear results, which assumes $t_c$ is a
% constant, we need to make further 
% approximations. Specifically, we consider vertically-isothermal disks
% and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
% H$, where $\Sigma$ is the surface density at the fiducial
% radius of interest. For example, the dimensionless diffusion
% coefficient becomes 
% \begin{align}\label{diff_coeff_dimensionless}
%   \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
%     c_v\Omega_k}. 
% \end{align}
% Furthermore, we will choose a temperature-dependent opacity model for
% simplicity, so that $\kappa_d=\kappa_d(T)$.
dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{1}{\hat{\eta}}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that the thermal relaxation time depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane. %The latter is due to the density-dependence in the diffusion time. 
%is then a constant for vertically-isothermal disks, 
%and we can directly apply previous results.

@@

% The physical thermal relaxation timescale $\beta$ is height-dependent,
% whilst for analytical discussion we assumed it is
% constant. Nevertheless,
It is instructive to compare the above thermal timescales
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the fiducial disk model,    
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is needed for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
% becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for three
opacity scales: $\hat{\kappa}_d=0.1, \,1 $ and $10$ at two heights:
$z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in each panel.

@@

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but are also
raised. In fact, lowering the opacity by an order of magnitude renders 
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the fundamental mode is
restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it develop in the inner disk, would only occur on very small
scales. However, our linear calculations suggest growth rates
eventually decay for large $\khat$.

Furthermore, modes with large $\khat$ may be subject to viscous decay
in a real disk. The viscous time for a perturbation lengthscale
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha_s\khat^2\Omega $, where
$\alpha_s$ is the standard alpha viscosity parameter
\citep{shakura73}. Setting $t_\mathrm{visc}$ to be longer than the
typical VSI growth timescale $1/\epsilon\Omega$ places an upper limit
$\khat \lesssim 30$ in a disk with $\alpha_s\sim 10^{-4}$ and
$\epsilon \sim 0.1$.

@@

\subsection{VSI in the MMSN}
Here we solve the linear problem in the MMSN with the scale and
height dependent thermal relaxation function defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $1\lesssim \khat \lesssim 30$ we either only find the
fundamental mode, or that it is the most unstable. The associated 
growth rates are also insensitive to $z_\mathrm{max}$. However, for
large $\khat$ and/or $z_\mathrm{max}$, e.g. $\khat=50$ or 
$z_\mathrm{max}=10H$, we also found modes with shorter growth times 
than those shown in Fig. \ref{mmsn_overall}, but these disturbances are 
concentrated near the vertical boundaries
\citep{nelson13,mcnally14}. Such modes may not be robust because    
they are affected by boundary conditions. % , and high $\khat$ modes are
% subject to viscous decay in a real disk.

We therefore take Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see that growth timescales are $\sim
30$---40 orbits, and the VSI operates from a few AU to tens of
AU. Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically
thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
mid-plane thermal relaxation timescales shown in
Fig. \ref{mmsn_bcrit_bcool}. For example, for $\khat=10$ we find 
$\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim 10$AU at $z=0$; and
Fig. \ref{mmsn_overall} show stabilization inside $\sim 8$AU. % The mode
% operates interior to the expected minimum radius,

% probably because the disk bcomes optically thin away from the
% midplane.

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ only for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with 
significantly reduced growth rates (when scaled by the local
rotation).  
%: Fig. \ref{mmsn_overall}) show that
%the growth time at

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer than the case with $\beta < \beta_\mathrm{crit}$ for
all $z$: we find $t_\mathrm{grow}\sim 40P_\mathrm{orb}$ at $r=50$AU,
but $t_\mathrm{grow}\sim 400P_\mathrm{orb}$ at $r=5$AU.

% the case with $\beta < \beta_\mathrm{crit}$ for all $z$.
